Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia

### 2024-08-18 manuscript/manuscript.Rmd
started to update the manuscript.

date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \usepackage[utf8]{inputenc}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: frontiers.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of "Large Cutting Tools" (LCTs) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was "at least partly under genetic control" [@corbey2016].
Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design.
Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of resharpening [@iovita2011; @mcpherron2000] whereas others find form to be unrelated to reduction intensity and more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shipton2015; @shipton2020].
Such debates about shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions including the origins of human culture [@corbey2016; @tennie2017; @shipton2015], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
To examine these questions, we studied the complete collection of Early Acheulean flaked pieces from 5 sites at Gona Project Area and compared them with Oldowan cores from 2 published sites at Gona.
By comparing shape variation to measures of flaking intensity and patterning, we sought to identify technological patterns that might reveal intent.

@@

Li, et al. [-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by an observed correlation (r=0.424) with SDI.
However, they also explain that "flaked area does not necessarily relate to the number of flake scars\...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage." (p. 6).
We suggest that what FAI actually captures is the spatial extent modifications to the surface of a core.
It is thus complementary to the measure of volume reduction provided by SDI and provides additional information to inform technological interpretations.
For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on "least-effort" shape imposition whereas the opposite pattern would be consistent with relatively intense resharpening of spatially restricted areas on the core.
A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly "designed" form achieved through extensive morphological and volumetric transformation. (**Figure** \@ref(fig:FAISDI))
In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.

@@

```{r fig2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Gona site map."}
knitr::include_graphics(here::here("figure/site map.png"))
```

@@

A PCA on GM transformed linear measures of the 33 artifacts identified two PCs accounting for 75.6% of the variance.
Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
This closely parallels the length vs. width and thickness tradeoff captured by 3DGM-PC1 (**Figure** \@ref(fig:fig3d)) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p \< 0.001, **Figure** \@ref(fig:figcomparison)A).
A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance. This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs. thick and pointed; **Table** \@ref(tab:tab1)) whereas 3DGM-PC2 more purely isolates convergence (**Figure** \@ref(fig:fig3d)). The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs 3 through 5, which comprise the contribution of the left and right lateral margins to relative thickness. Use of high-resolution, coordinate-based scan data thus generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by the linear and 3D analyses remains similar. Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed. A stepwise regression ($r^2$=0.625, F(4,28)=11.697, p<0.001, Probability-of-F-to-enter <= 0.050; Probability-of-F-to-remove >= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.

@@

```{r figcomparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Comparison between linear measurement and 3d measurement."}
knitr::include_graphics(here::here("figure/Fig.3D comparison.png"))
```

@@

Visual estimation of flaked area approximated 3DGM measurement very closely (r2=0.932, p < 0.001, $\beta$ = 1.051; **Figure** \@ref(fig:figcomparison)C) and without any systematic bias (paired t-test: mean difference = -0.015%, 95% CI = 2.02% to 1.99%, p = 0.987). Individual errors ranged between -10.46% and 14.24%. We thus considered visual estimation to be reliable in our sample.

>>>>

date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \usepackage[utf8]{inputenc}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: jhe.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of “Large Cutting Tools” (LCTs; cf. Long Core Tools [Shea, -@shea2013]) in the Early Acheulean [@isaac1976; @kuhn2020; @holloway1969].
However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was “at least partly under genetic control” [@corbey2016: 6].
Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design.
Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of function-driven resharpening [@iovita2011; @mcpherron2000] whereas others find it more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shipton2015; @shipton2020].
Such debates about the shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions, including the origins of human culture [@corbey2016; @liu2023; @tennie2017; @shipton2015], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
To examine these questions, we studied the complete collection of Early Acheulean (ca. 1.7 –1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared them with all of the flaked pieces from two published Oldowan (\>2.5 Ma) sites at Gona (OGS7 and EG10; Semaw, 2000; Semaw et al., 2003; Stout et al., 2010).
By comparing variation in overall artifact shape (or “Bauplan” [Lycett and Gowlett, 2008)) to measures of flaking intensity and coverage, we sought to identify technological patterns that might reveal intent.
The possibilities we considered include: a) intentional shape standardization reflecting normative expectations of what handaxes should look like, b) regularities of overall form arising as by-products of the intentional pursuit of particular morpho-functional attributes, and c) the absence of any intentional form imposition as in the case of simple debitage.

@@

Li, et al. [-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by an observed correlation (r=0.424) with SDI.
However, they also explain that "flaked area does not necessarily relate to the number of flake scars...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage." (p. 6).
We suggest that what FAI actually captures is the spatial extent modifications to the surface of a core.
It is thus complementary to the measure of volume reduction provided by SDI and provides additional information to inform technological interpretations.
For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on "least-effort" shape imposition whereas the opposite pattern would be consistent with relatively intense resharpening of spatially restricted areas on the core.
A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly "designed" form achieved through extensive morphological and volumetric transformation.
(**Figure** \@ref(fig:FAISDI)) In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.

@@

<!-- ```{r fig2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Gona site map."} -->

<!-- knitr::include_graphics(here::here("figure/site map.png")) -->

<!-- ``` -->

@@

A PCA on GM transformed linear measures of the 33 artifacts identified two PCs accounting for 75.6% of the variance.
Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
This closely parallels the length vs. width and thickness tradeoff captured by 3DGM-PC1 (**Figure** \@ref(fig:fig3d)) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p \< 0.001, **Figure** \@ref(fig:figcomparison)A).
A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance.
This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs. thick and pointed; **Table** \@ref(tab:tab1)) whereas 3DGM-PC2 more purely isolates convergence (**Figure** \@ref(fig:fig3d)).
The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs 3 through 5, which comprise the contribution of the left and right lateral margins to relative thickness.
Use of high-resolution, coordinate-based scan data thus generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by the linear and 3D analyses remains similar.
Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.
A stepwise regression ($r^2$=0.625, F(4,28)=11.697, p\<0.001, Probability-of-F-to-enter \<= 0.050; Probability-of-F-to-remove \>= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.

@@

```{r figcomparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Comparison between linear measurement and 3d measurement."}
knitr::include_graphics(here::here("figure/Fig.3D comparison new1.png"))
```

@@

Visual estimation of flaked area approximated 3DGM measurement very closely (r2=0.932, p \< 0.001, $\beta$ = 1.051; **Figure** \@ref(fig:figcomparison)C) and without any systematic bias (paired t-test: mean difference = -0.015%, 95% CI = 2.02% to 1.99%, p = 0.987).
Individual errors ranged between -10.46% and 14.24%.
We thus considered visual estimation to be reliable in our sample.
